Mechanising Hylomorphisms for Extraction of Executable Code

### 2025-03-24 paper/paper.tex
Fixes

Hylomorphisms are highly versatile, non-structural recursive algorithms that are
capable of implementing structural recursion as well as other known recursion
schemes like mutual recursion, accumulators, primitive recursios, and
course-of-values iteration (dynamic programming)~\cite{HinzeWG15}. However, this
versatility comes with a trade-off: because hylomorphisms are not inherently
structurally recursive, special attention is required when encoding them in a
type theory. Specifically, the ``divide'' phase of the algorithm must be
recursive, meaning the input needs to be broken down into smaller parts, which
do not necessarily align with the structure of the input data. On the other
hand, a key benefit is the reusability of termination proofs: once the
correctness of the divide phase is established, the same proof can be reused for
any other conquer phase we choose. Moreover, in the case of structural
recursion, the divide phase naturally follows the structure of the data, making
it correct by definition and eliminating the need for additional proofs.

@@

Instead, we prove the acid rain theorem of the specific parametric function
\coq{tau}. This essentially is a recursive function on lists which uses the
algebra for both the recursive step and for the base case where it is used in
the hylomorphism to continue the recursion:
\begin{coqcode}
Definition tau (l : list A) (a : Alg (ListF A) B) : App (ListF A) B -> B :=
  fun x => match x with | MkCont sx kx => match sx with
  | s_nil => fun _ => (hylo a ilist_coalg) l
  | s_cons h => fun kx => a (MkCont (s_cons h) kx)
  end kx end.
\end{coqcode}
Our acid rain theorem is stated as follows:
\begin{coqcode}
  hylo a l_out \o hylo (tau l l_in) c =e hylo (tau l a) c
\end{coqcode}

@@

The problem of \emph{nonstructural recursion} (including
divide-and-conquer algorithms) is well-studied~\cite{BoveKS16}.
% A distinction between standard and
% nonstandard recursion lies in the fact that standard recursion adheres strictly
% to recursing on pattern variables, avoiding reliance on results from other
% function calls. Nonstandard recursion may deviate from this convention, allowing
% for more flexibility.
Certain functions that are not structurally recursive can be reformulated using
a nonstandard approach to achieve structural recursion.  For exammple, the
mergesort in Coq's standard library uses an ``explicit stack of pending merges''
in order to avoid issues with nonstructural definitions. There is a major
downside, however; as noted by Abreu et al., the result is ``barely recognisable
as a form of mergesort''~\cite{AbreuDHJMS23}.  There are common approaches in
Coq to deal with termination~\cite{DBLP:journals/pacmpl/SozeauM19,
  DBLP:conf/icfp/Sozeau07}, but none of these approaches address program
calculation techniques, and the mechanisation of fusion laws.

% \textbf{Fusion Optimisation}.
% %
% \label{sec:fusion}
The fusion laws that we formalise in this paper are applied in the more general
context of compilers for functional programming lanmguages.  Fusion has been
applied to recursive tree traversals, in order to reduce the number of times
each section of the tree must be visited. Research has demonstrated that this
sort of fusion can be performed automatically by a
compiler~\cite{10.1145/3133900}. The correctness of tree traversal fusion has
been verified mechanically in Coq~\cite{10.1145/3377555.3377884}, but the
authors do not formalise unfolds and hylomorphisms.

>>>>

Hylomorphisms are highly versatile, non-structural recursive algorithms that are
capable of implementing structural recursion as well as other known recursion
schemes like mutual recursion, accumulators, primitive recursion, and
course-of-values iteration (dynamic programming)~\cite{HinzeWG15}. However, this
versatility comes with a trade-off: because hylomorphisms are not inherently
structurally recursive, special attention is required when encoding them in a
type theory. Specifically, the ``divide'' phase of the algorithm must be
recursive, meaning the input needs to be broken down into smaller parts, which
do not necessarily align with the structure of the input data. On the other
hand, a key benefit is the reusability of termination proofs: once the
correctness of the divide phase is established, the same proof can be reused for
any other conquer phase we choose. Moreover, in the case of structural
recursion, the divide phase naturally follows the structure of the data, making
it correct by definition and eliminating the need for additional proofs.

@@

We prove a specialised version of the acid rain theorem that fuses
hylomorphisms defined in terms of the following function \coq{tau} that maps
list algebras to list algebras.  This essentially is a function on lists which
uses the algebra for both the recursive step and for the base case where it is
used in the hylomorphism to continue the recursion:
\begin{coqcode}
Definition tau (l : list A) (a : Alg (ListF A) B) : App (ListF A) B -> B :=
  fun x => match x with | MkCont sx kx => match sx with
  | s_nil => fun _ => (hylo a ilist_coalg) l
  | s_cons h => fun kx => a (MkCont (s_cons h) kx)
  end kx end.
\end{coqcode}
Our acid rain theorem is stated as follows:
\begin{coqcode}
  hylo a l_out \o hylo (tau l l_in) c =e hylo (tau l a) c
\end{coqcode}

@@

The problem of \emph{nonstructural recursion} (including
divide-and-conquer algorithms) is well-studied~\cite{BoveKS16}.
% A distinction between standard and
% nonstandard recursion lies in the fact that standard recursion adheres strictly
% to recursing on pattern variables, avoiding reliance on results from other
% function calls. Nonstandard recursion may deviate from this convention, allowing
% for more flexibility.
Certain functions that are not structurally recursive can be reformulated using
a nonstandard approach to achieve structural recursion.  For example, the
mergesort in Coq's standard library uses an ``explicit stack of pending merges''
in order to avoid issues with nonstructural definitions. There is a major
downside, however; as noted by Abreu et al., the result is ``barely recognisable
as a form of mergesort''~\cite{AbreuDHJMS23}.  There are common approaches in
Coq to deal with termination~\cite{DBLP:journals/pacmpl/SozeauM19,
  DBLP:conf/icfp/Sozeau07}, but none of these approaches address program
calculation techniques, and the mechanisation of fusion laws.

% \textbf{Fusion Optimisation}.
% %
% \label{sec:fusion}
The fusion laws that we formalise in this paper are applied in the more general
context of compilers for functional programming languages.  Fusion has been
applied to recursive tree traversals, in order to reduce the number of times
each section of the tree must be visited. Research has demonstrated that this
sort of fusion can be performed automatically by a
compiler~\cite{10.1145/3133900}. The correctness of tree traversal fusion has
been verified mechanically in Coq~\cite{10.1145/3377555.3377884}, but the
authors do not formalise unfolds and hylomorphisms.
